---
title: "Compare to the ACS data"
author: "Nate Lant"
date: "2/19/2020"
output: 
  html_document:
    code_folding: hide
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# install.packages("devtools")
# devtools::install_github("byu-transpolab/nhts2017")
library(nhts2017)
library(tidyverse)
library(scales)
library(pander)

# for help understanding the kableExtra library 
# (https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
# there may be some issues with this table going to html vs pdf and latex
library(kableExtra)
library(Stack)

source("R/functions.R")

my_nhts <- read_rds("data/nhts_data.rds")
my_persons <- read_rds("data/persons.rds")
```

## American Community Survey
The ACS provides information on disability status and can be used to validate fidings from the NHTS. 
The tables can be found [here.](https://data.census.gov/cedsci/table?q=S18&d=ACS%201-Year%20Estimates%20Subject%20Tables&tid=ACSST1Y2018.S1810&vintage=2018)

## NHTS Data on Disabilities
Different from the previous page, this table of distributions only compares the abled population and the
disabled population. This table also includes an "Overall" column which shows the distribution of the total
civilian population.

The variables and grouping of variables are designed to match the ACS variables (i.e. age, mode, education)

```{r mutate.data}
# people data for employment and education comparison
# create column for just abled and disabled
data <- nhts_persons %>% 
  filter(r_age > 16) %>%
  mutate(status = case_when( medcond == "01" ~ "Disabled", 
                             medcond == "02" ~ "Abled",
                             TRUE ~ "NA"),
         Employment = case_when( worker == "01" ~ "Employed",
                                 worker == "02" ~ "Not in Labor Force",
                                 TRUE ~ "NA"),
         Education = case_when( educ == "01" ~ "Less than high school graduate",
                           educ == "02" ~ "High school graduate (or equivalent)",
                           educ == "03" ~ "Some college or associates degree",
                           educ == "04" | educ == "05" ~ "Bachelor's degree or higher",
                           TRUE ~ "NA"),
         Education = fct_relevel(Education, "Less than high school graduate",
                                 "High school graduate (or equivalent)", 
                                 "Some college or associates degree",
                                 "Bachelor's degree or higher")
         ) %>%
  filter(status != "NA", 
         Employment != "NA",
         Education != "NA") 

# trips data for mode choice
trips_data <- my_nhts %>%
  filter(r_age > 16,
         whyto == "03") %>%
  mutate(status = case_when( medcond == "01" ~ "Disabled", 
                             medcond == "02" ~ "Abled",
                             TRUE ~ "NA"),
         Employment = case_when( worker == "01" ~ "Employed",
                                 worker == "02" ~ "Not in Labor Force",
                                 TRUE ~ "NA"),
         Commute = case_when(wrk_home == "01" ~ "Work from Home",
           
                             psgr_flg == "02" & trptrans == "03" ~ "Car/Van/Truck (Driver)",
                             psgr_flg == "02" & trptrans == "04" ~ "Car/Van/Truck (Driver)",
                             psgr_flg == "02" & trptrans == "05" ~ "Car/Van/Truck (Driver)",
                             psgr_flg == "02" & trptrans == "06" ~ "Car/Van/Truck (Driver)",
                             # 
                             psgr_flg == "01" & trptrans == "03" ~ "Car/Van/Truck (Passenger)",
                             psgr_flg == "01" & trptrans == "04" ~ "Car/Van/Truck (Passenger)",
                             psgr_flg == "01" & trptrans == "05" ~ "Car/Van/Truck (Passenger)",
                             psgr_flg == "01" & trptrans == "06" ~ "Car/Van/Truck (Passenger)",
                            
                             trptrans == "01" ~ "Walk",
                             
                             trptrans == "11" | trptrans == "16"  ~ "Local Transit",
                             
                             TRUE  ~ "Taxi/Motorcycle/Bike or Other"
                     )) %>%
  filter(status != "NA")
  
```

```{r function.data}

# Build funciton to create distribution table that includes the total distribution

build_table <- function(data, variable){
  
  quote_var <- enquo(variable)
  
  # calulate statistics from the data
data %>%
  group_by(status, !!quote_var) %>%
  summarise(population = sum(wtperfin)) %>%
  mutate(distribution = percent(population/sum(population), accuracy = 0.1),
         Variable = as.factor(!!quote_var)) %>%
  select(status, Variable, distribution) %>%
  spread(status, distribution) %>%
  
  left_join(
    data %>% 
      group_by(!!quote_var) %>%
      summarise(population = sum(wtperfin)) %>%
      mutate(Overall = percent(population/sum(population), accuracy = 0.1),
             Variable = as.factor(!!quote_var)) %>%
      select(-population, -!!quote_var),
    by = c("Variable")
  )
}
```

## Compare ACS with NHTS
Note: in comparing mode choice, the NHTS distribution is of total work trips; I don't know how ACS calculates "Commute
to work".
```{r comparison.table, warning=F, message=F}
# Building the NHTS distribution table
emp <- build_table(data, Employment) 
#this is percentage of total trips
comm <- build_table(trips_data, Commute) 

edu <- build_table(data, Education) 

Stack(emp, comm) %>% Stack(edu) %>%
  
  
  # pipe into the kable function
  kable(caption = "2017 NHTS Data Distribution on Ability") %>%
  kable_styling("striped", full_width = F) %>%
  # Specify which row numbers are grouped together (i.e. from 1, to 10)
  pack_rows("Employment", 1, 2) %>%
  pack_rows("Commuting to Work", 3, 8) %>%
  pack_rows("Education", 9, 12) 

```

```{r acs, message=F, warning=F}

#=======================================================================

# building the ACS distribution
emp_acs <- tibble(
  Variable = c("Employed", "Not in Labor Force"),
  ACS_Abled = percent(c(0.673, 0.295)),
  ACS_Disabled = percent(c(0.24, 0.762)),
  ACS_Overall = percent(c(0.609, 0.359))
)

comm_acs <- tibble(
  Variable = c("Car/Van/Truck (Driver)", "Car/Van/Truck (Passenger)", "Local Transit", 
               "Taxi/Motorcycle/Bike or Other", "Walk", "Work from Home"),
  Abled = percent(c(0.767, 0.088, 0.049, 0.025, 0.018, 0.053)),
  Disabled = percent(c(.711, .116, .051, .033, .027, .061)),
  Overall = percent(c(0.763, 0.09, 0.049, 0.025, 0.019, 0.053))
)

edu_acs <- tibble(
  Variable = c("Less than high school graduate", "High school graduate (or equivalent)", 
               "Some college or associates degree", "Bachelor's degree or higher"),
  Abled = percent(c(0.098, 0.253, 0.29, .359)),
  Disabled = percent(c(0.20, 0.338, .284, .178)),
  Overall = percent(c(.115, 0.267, 0.289, .33))
)

Stack(emp_acs, comm_acs) %>% Stack(edu_acs)
  
  
  # pipe into the kable function
  kable(caption = "2010-2018 ACS Data Distribution on Ability") %>%
  kable_styling("striped", full_width = F) %>%
  # Specify which row numbers are grouped together (i.e. from 1, to 10)
  pack_rows("Employment", 1, 2) %>%
  pack_rows("Commuting to Work", 3, 8) %>%
  pack_rows("Education", 9, 12) 


```

## NHTS Data including Wheelchair and Part-time Employment
