---
title: "DAP Utility Function - Multinomial Logit Model"
author: "Nate Lant"
date: "2/26/2020"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(nhts2017)
library(tidyverse)
```

## Data Organization
The list of persons will be filtered by person type.
* (FW) Full-time worker
* (PW) Part-time worker
* (US) University Student - Claims their educational attainment is higher than high school degree and confirmed their student status as yes.
* (NW) Non-working adult
* (RT) Retired
* (SD) Driving age schoolchild
* (SP) Pre-driving age schoolchild
* (PS) Pre-school child
* *(WW) Worker with wheelchair* - responded they use a motorized scooter or a manual or mechanical wheelchair and that they are a worker.
* *(WN) Non-worker with wheelchair*

### Labeling Each Person
Each person needs to have the attributes age, gender, auto ownership, and income.

```{r create.data, echo=F}
my_persons <- nhts_persons %>%
  mutate(wheelchair_status = 
           case_when(
             #wheelchair or non-wheelchair status
             w_chair == "07" | w_mtrchr == "08" | w_scootr == "06" ~ T,
             TRUE ~ F
             ),
         #broken....
         university_status = 
           case_when(
             educ == "03" & schtyp == "01" ~ T,
             educ == "04" & schtyp == "01" ~ T,
             educ == "05" & schtyp == "01" ~ T,
             TRUE ~ F
           ),
         person_type = 
           case_when(
             wheelchair_status == T & worker == "01" ~ "WW",
             wheelchair_status == T & worker == "02" ~ "WN",
             university_status == T ~ "US",
             wkftpt == "01" ~ "FW",
             wkftpt == "02" ~ "PW",
             prmact == "06" ~ "RT",
             r_age >= 18 & worker == "02" ~ "NW", 
             r_age < 18 & r_age >= 16 & schtyp == "01" ~ "SD",
             r_age < 16 & r_age >= 5 & schtyp == "01" ~ "SP",
             r_age < 5 ~ "PS"
             ),
         # relevel the factors
         person_type = fct_relevel(person_type, 
                                   "FW", "PW", "US", "NW", "RT", "SD", "SP", "PS", "WW", "WN")
           ) 

my_persons %>%
  group_by(person_type) %>%
  summarise(count = n())
         

```
