---
title: "Trips, Activities, and Tours"
author: "Nate Lant"
date: "2/7/2020"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## FUNCTION READ IN FROM 

library(tidyverse)
library(nhts2017)
library(scales)
library(pander)
library(lubridate)
source("R/functions.R")

# for help understanding the kableExtra library 
# (https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
# there may be some issues with this table going to html vs pdf and latex
library(kableExtra)
library(Stack)

# read in the trips file
trips_edited <- read_rds("data/nhts_data.rds") %>%
  filter(r_age > 18,
         r_age < 65, 
         msasize == "04")
persons_edited <- read_rds("data/persons.rds") %>%
  filter(r_age > 18,
         r_age < 65, 
         msasize == "04")
# already filters people that are in regular sized cities
myactivities <- read_rds("data/activities_msa.rds")
mytours <- read_rds("data/tours.rds")
```



# Trips
The NHTS survey collects data on each person about every trip they take (note that some people did not 
take any trips). In this data set, we will filter the number of trips by 

* people ages 18-65

* MSA size 1,000,000 - 2,999,999

```{r trips.data, message=F, warning=F}
# trips edited comes from joining the persons file with trips. Includes individuals with 0 trips
trips_print <- trips_edited %>%
  select(houseid, personid, strttime, endtime, whyfrom, whyto)

trips_print %>% 
  head(n = 5) %>% 
  formattable::formattable()

```

# Activities
Each trip goes from one activity to the next. By manipulating the data set, we can create a list of
activities based on the purpose and duration of each trip using `build_activities`.

```{r activities.data, message=F, warning=F}
activities_print <- trips_print %>% 
  filter(houseid %in% c("30000301", "30000309", "30000312")) %>%
  build_activities()

activities_print %>% 
  head(n= 5) %>% 
  formattable::formattable()
```

# Activities with Tour Attributes
Using the function `add_tour()` each activity can be assigned to a daily activity pattern (DAP) which 
describes the type of tour or group of tours a person makes during the day.
A tour is a collection of activities from when you leave home to when you return home. Often, people 
take more than one tour per day. Tours are classified as Mandatory (work or school), Non-Mandatory, 
and Home. For the purpose here, we label each tour as "W" for work, "S" for school, and "NM" for
non-mandatory (everything else). Home trips will show up as NA for now.

```{r tours.data, message=F, warning=F}
myactivities %>% 
  select(houseid, personid, activity, activity_number, arrive, 
         depart, tour_count, tour_class, DAP) %>%
  head(n=5) %>% 
  formattable::formattable()
```

# Tours
Using the `build_tours()` function, a list of activities can be converted into tours--which is the same list of persons.
```{r tours, message=F, warning=F}
mytours %>%
  head(n=5) %>% 
  formattable::formattable()
```


## Distribution of DAP
What kinds of daily activity patterns (DAP) are being made by people with disabilities?

```{r distribution.tourclass,  warning=F, message=F}
# Distribution table (percent of persons)
mytours %>% 
  distributions(DAP) %>% 
  pander(caption = "Daily Tours (List of Persons)")

myactivities %>% 
  distributions(DAP) %>% 
  pander(caption = "Daily Activities (List of Activities)")

```

## Estimated Population and Counts of DAP Groups

This table shows the same distribution but of total counts *from the list of persons NOT activities*. 
The total number of persons is 23,906 in 
this set. This data has been filtered by age and by MSA size, as described previously. Using the 
weighted population estimates in the NHTS data, the probable total population has been summed in each group.

```{r counts, warning=F, message=F}
# this calculates the population in each group, based on the tour distribution
mytours %>%
  filter(Ability != "NA", 
         DAP != "NA") %>%
    # Calculate number of people in each combination of dap and ability.
    group_by(Ability, DAP) %>%
    summarise(population = sum(wtperfin)) %>%
    spread(Ability, population) %>%
  pander(caption = "Population of Each DAP Group")

# this calculates the frequency of each DAP from each group
mytours %>%
  filter(Ability != "NA", 
         DAP != "NA") %>%
    # Calculate number of people in each combination of dap and ability.
    group_by(Ability, DAP) %>%
    summarise(Frequency = n()) %>%
    spread(Ability, Frequency) %>%
  pander(caption = "Frequency of DAPs from NHTS Survey Respondents")

```






