---
title: "Trips, Activities, and Tours"
author: "Nate Lant"
date: "2/7/2020"
output:
  html_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

## FUNCTION READ IN FROM 

library(tidyverse)
library(nhts2017)
library(scales)
library(pander)
library(lubridate)
source("R/functions.R")

# for help understanding the kableExtra library 
# (https://cran.r-project.org/web/packages/kableExtra/vignettes/awesome_table_in_html.html)
# there may be some issues with this table going to html vs pdf and latex
library(kableExtra)
library(Stack)

# read in the trips file
trips_edited <- read_rds("data/nhts_data.rds") %>%
  filter(r_age > 18,
         r_age < 65, 
         msasize == "04")
persons_edited <- read_rds("data/persons.rds") %>%
  filter(r_age > 18,
         r_age < 65, 
         msasize == "04")
# already filters people that are in regular sized cities
activities <- read_rds("data/activities_msa.rds")
```



# Trips
The NHTS survey collects data on each person about every trip they take (note that some people did not 
take any trips). In this data set, we will filter the number of trips by 

* people ages 18-65

* MSA size 1,000,000 - 2,999,999

```{r trips.data, echo=F, message=F, warning=F}
# trips edited comes from joining the persons file with trips. Includes individuals with 0 trips
trips_print <- trips_edited %>%
  select(houseid, personid, strttime, endtime, whyfrom, whyto)

trips_print %>% 
  head(n = 5) %>% 
  formattable::formattable()

```

# Activities
Each trip goes from one activity to the next. By manipulating the data set, we can create a list of
activities based on the purpose and duration of each trip using `build_activities`.

```{r activities.data, echo=F, message=F, warning=F}
activities_print <- trips_print %>% 
  filter(houseid %in% c(30000301, 30000309, 30000312)) %>%
  build_activities()

activities_print %>% 
  head(n= 5) %>% 
  formattable::formattable()
```

# Activities with Tour Attributes
Using the function `add_tour()` each activity can be assigned to a daily activity pattern (DAP) which 
describes the type of tour or group of tours a person makes during the day.
A tour is a collection of activities from when you leave home to when you return home. Often, people 
take more than one tour per day. Tours are classified as Mandatory (work or school), Non-Mandatory, 
and Home. For the purpose here, we label each tour as "W" for work, "S" for school, and "NM" for
non-mandatory (everything else). Home trips will show up as NA for now.

```{r tours.data, echo=F, message=F, warning=F}
activities %>% 
  select(houseid, personid, activity, activity_number, arrive, 
         depart, tour_count, tour_class, DAP) %>%
  head(n=5) %>% 
  formattable::formattable()
```


## Distribution of DAP
What kinds of daily activity patterns (DAP) are being made by people with disabilities?

```{r distribution.tourclass, echo=F, warning=F, message=F}
# Distribution table (percent of activities not persons)
activities %>% 
  distributions(DAP) %>% 
  pander()

```






